Copulas and their potential for ecology

### 2020-02-03 Paper.Rmd
correction for proof started

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables. Some of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. C and D show normalized ranks of A, B, 
respectively. (E, F) Reflections
of C, D, respectively, about a vertical axis. Subscripts h and v 
in the axis labels stand for "horizontal" and "vertical".\label{fig_cop_pedag1}}
\end{center}
\end{figure}

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the association 
between two or more variables that is mathematically provably 
complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*. Copula approaches separate information in a bivariate 
random variable into two parts: the information in the 
marginal distributions (which says nothing about the association between the 
variables), and the remaining information (which is solely about
the association). We here introduce some simple concepts, based on ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
formally in the "Background" section below (section \ref{Background}). Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the association
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cumulative distribution function of the $x_i$ and $y_i$,
respectively. We henceforth abbreviate "cumulative distribution function" as "cdf". 
See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described 
(section \ref{Background}). Note that Pearson correlation, even though it is the most commonly
used measure of association, is modified by monotonic
transformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals [@Genest2007]. 
Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the variables. 
They provide information solely on the dependence between those variables.

A main benefit of a copula approach is 
that it can detect associations in the tails of distributions, and asymmetries of such associations. 
Tail association (introduced formally in section 
\ref{Background}) is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail association than right- or upper-tail association
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated 
(Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail associations 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see sections on Background and
methods for Q1 below, sections \ref{Background} and \ref{Methods}). 
Although copulas can be used to model the entire
complex dependence structure of data, we have found tail associations to
be a useful component of that structure, so tail association 
is a main focus of this paper. We give four examples below which
illustrate some of the ecological meaning of tail associations.

The goal of this paper is to explore the potential for 
applications of copulas in ecology and to 
estimate to what extent ecological understanding may benefit from
using copulas. We provide an introduction
to copula concepts in section \ref{Background}. 
Readers can find additional abbreviated [@Anderson2018; @Genest2007] 
and comprehensive [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]
introductions elsewhere. Copulas have already been used effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2018], but such usage is
still rare. Whereas @Anderson2018 provide a specific method, based
on copulas, for important problems of the 
analysis of multivariate population count data,
we instead advocate more broadly for the application of copulas across ecology.
Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate complex dependence structures, including asymmetric tail 
associations, and that greater use of copulas can illuminate underlying processes.
We believe copula approaches are among the tools all ecologists 
should be considering for analysis of their data in the 21st century.

@@

A third reason why we may expect an understanding of complex 
dependence structures and tail associations to benefit 
ecology is that such an approach may help illuminate 
causal mechanisms between variables. If two species,
Sp1 and Sp2, are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest Sp2
is the dominant competitor. Other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on associations. However, 
examination of tail associations may
be combined with biological reasoning to rule out
some causal hypotheses which could not previously be eliminated. 
@Popovic2018 used copulas to illuminate causal relationships
between species.

@@

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do positively associated 
ecological variables show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from a standard or default case. 
(Q2) If the answer to
Q1 is "yes", then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3), what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After section \ref{Background}, 'Background on copulas', we address Q1 
(sections \ref{Data} through \ref{Results} of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sections \ref{Causes} through \ref{Results_Q2}), using 
simple models. We address Q3 (sections \ref{Consequences} through \ref{Results_Q3}) using 
both data and models. Section \ref{Discussion} is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagrammatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have also been used effectively, but rarely so far, in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider use of
copula descriptions of dependence in ecology will be substantial.

@@

We introduce some standard copula families, as examples, and also
because we will fit these families to data. The bivariate normal
copula family, which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of association
between the variables. A normal copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cdf of a (univariate) standard 
normal and $\Phi_{2,p}$ is the cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. Formulas for all the 
copula families we use were provided by @Brechmann2013. 
However, formulas for copulas are often not as 
directly informative about copula properties as probability density function (pdf) 
pictures; Fig. \ref{figped_N} has 
example pdfs corresponding to bivariate normal copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
associations between the two variables in the left and right tails. 
The Clayton copula family,
of which we already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to normal copulas, Clayton copulas
have stronger left- than right-tail association. The formula is 
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). Higher values of the parameter,
$p$ produce copulas with higher Kendall or Spearman correlation. The survival
Clayton family is the symmetric opposite of the Clayton family, showing
stronger right- than left tail association (Fig. \ref{SM-figped_SC}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail association, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula, @joe2014_dependence and 
@Brechmann2013 for further
information on these and other families.

@@

Plots using normalized ranks were proposed in the Introduction as being conceptually
similar to copulas; we explain the connection. As stated above, the copula associated
with a random variable $(X,Y)$ is the cdf of the random variable
$(F_X(X),F_Y(Y))$. Given a sample $(x_i,y_i)$, $i=1,\ldots,n$, from $(X,Y)$, let $\hat{F}_X$ and
$\hat{F}_Y$ be the empirical cdfs associated with the $x_i$ and $y_i$, respectively.
These are step functions of $x$ and $y$, respectively, that start at $0$ for low values of 
$x$ and $y$ and jump by $1/n$
at each of the data points. Therefore $\hat{F}_X(x_j)$ is the rank of $x_j$ in
the set $\{x_i : i=1,\ldots,n\}$, divided by $n$, 
and analogously $\hat{F}_Y(y_j)$ is the rank of $y_j$ in the set 
$\{y_i : i=1,\dots,n\}$, divided
by $n$. For large $n$, $\hat{F}_X(x_i)$ and $\hat{F}_Y(y_i)$ 
approximate the cdfs $F_X$ and $F_Y$, respectively. 
But $\hat{F}_X(x_i)$ equals the normalized rank of $x_i$ times
$n/(n+1)$, and likewise $\hat{F}_Y(y_i)$ equals the normalized rank of $y_i$ times
$n/(n+1)$. So the normalized rank in turn approximates the empirical cdf and therefore 
the cdf. Thus the normalized rank pairs $(u_i,v_i)$ for $i=1,\ldots,n$ can be 
regarded as approximate samples from the random variable $(F_X(X),F_Y(Y))$, 
which is the random variable associated with the copula of $(X,Y)$.
The scatterplot of $v_i$ versus $u_i$ can be used to infer aspects of copula 
structure (see section \ref{Methods} below).

@@

The datasets we used included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data (Table \ref{tab_data_info}), 
selected to span multiple fields and levels
of organization within ecology. Copula structure 
between atmospheric weather variables such as rainfall or wind speed, 
as measured in multiple locations through time, has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [RaCA; @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (megagrams C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for locations).
Species-trait data were average species basal metabolic rate (BMR, KJ per hour) and 
body mass (grams) for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the association has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years. 
Our second population dataset comprised average annual plankton abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells.
These data were computed from the Continuous Plankton Recorder survey of the 
UK Marine Biological Survey.
Community-level data, obtained from the Cedar Creek Ecosystem Science Reserve, were 
plant aboveground biomass [@ccdataple120] and Shannon's diversity index (computed from plant 
species percent cover data [@ccdatapce120]) for `r nrow(d_cc[[5]])` plots [@ccdataplotse120], 
each 9m by 9m, as sampled in the years 1996-2000 and 2007, each year 
analyzed separately [@Tilman2001; @Tilman2006]. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. See Appendix \ref{SM-Data_details} for additional data details.

@@

<!--what we did and why is it important to study?-->
We showed that non-normal copula structure and asymmetric tail associations are 
common across multiple sub-disciplines
in ecology, although these facets of data are only occasionally 
accounted for [@Valpine2014; @Anderson2018; @Popovic2018]. We hypothesized 
mechanisms that may cause non-normal copula structures and asymmetric tail associations;
we discuss below how commonly some of our
mechanisms may operate. 
We also demonstrated important consequences of 
non-normal copula structure and asymmetric tail associations for ecology. 
For instance, the skewness
of a spatial-average time series is influenced by asymmetric tail associations between its 
constituent time series: predominantly right-tail-associated local time series 
can lead to "spiky" spatially averaged time series, with large outbreaks; and 
predominantly left-tail-associated local time series can lead to spatially
averaged time series showing accentuated "crashes".
Thus tail associations could have implications for pests and exploited species.
Extinction risk and Taylor's law can also be altered by tail association 
patterns across space. 
In our view, our results make it reasonable to suggest 
that a more comprehensive understanding 
of many ecological phenomena may be possible if a
complete copula characterization of associations between variables is employed. 
Copula statistics are
well developed [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017], 
and have been introduced in accessible formats [@Anderson2018; @Genest2007]. Furthermore, 
open-source computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in R). Ecologists can apply these tools
immediately. We created several interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that built upon existing copula methods.

@@

<!-- We have only given a few mechanisms, but there are likely to be others, and discoverig
additional mechanisms of copula structure is certainly worthy of future work. -->
Additional mechanisms of non-normal copula structure and asymmetric tail associations
probably also operate. For instance, 
measurement error may modify copula structure. 
Our models investigating potential causes of copula structure 
were intentionally simple in other respects, 
too, not including factors such as
delayed density dependence, dispersal, population stage structure, trophic
interactions, etc.; and we did not comprehensively 
explore parameter space for our models. We re-emphasize 
that fuller explorations, in future work,
of some of our models and of variant model may be informative. 
We hope by enumerating a few potential mechanisms of copula structure
we will inspire additional research on the
potentially numerous mechanisms that may operate in diverse datasets,
and their relative importance under different circumstances.

>>>>

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables. Some of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. (C) and (D) show normalized ranks of (A), (B), 
respectively. (E, F) Reflections
of (C), (D), respectively, about a vertical axis. Subscripts h and v 
in the axis labels stand for "horizontal" and "vertical".\label{fig_cop_pedag1}}
\end{center}
\end{figure}

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the association 
between two or more variables that is mathematically provably 
complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*. Copula approaches separate information in a bivariate 
random variable into two parts: the information in the 
marginal distributions (which says nothing about the association between the 
variables), and the remaining information (which is solely about
the association). We here introduce some simple concepts, based on ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
formally in the "Background" section below (section \ref{Background}). Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the association
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cumulative distribution function of the $x_i$ and $y_i$,
respectively. We henceforth abbreviate "cumulative distribution function" as "cdf". 
See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described 
(section \ref{Background}). Note that Pearson correlation, even though it is the most commonly
used measure of association, is modified by monotonic
transformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals [@Genest2007]. 
Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the variables. 
They provide information solely on the dependence between those variables.

A main benefit of a copula approach is 
that it can detect associations in the tails of distributions, and asymmetries of such associations. 
Tail association (introduced formally in section 
\ref{Background}) is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail association than right- or upper-tail association
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated 
(Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail associations 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see sections on Background and
methods for Q1 below, sections \ref{Background} and \ref{Methods}). 
Although copulas can be used to model the entire
complex dependence structure of data, we have found tail associations to
be a useful component of that structure, so tail association 
is a main focus of this paper. We give four examples, below, which
illustrate some of the ecological meaning of tail associations.

The goal of this paper is to explore the potential for 
applications of copulas in ecology and to 
estimate to what extent ecological understanding may benefit from
using copulas. We provide an introduction
to copula concepts in section \ref{Background}. 
Readers can find additional abbreviated [@Anderson2018; @Genest2007] 
and comprehensive [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]
introductions elsewhere. Copulas have already been used effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2019], but such usage is
still rare. Whereas @Anderson2018 provide a specific method, based
on copulas, for important problems of the 
analysis of multivariate population count data,
we instead advocate more broadly for the application of copulas across ecology.
Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate complex dependence structures, including asymmetric tail 
associations, and that greater use of copulas can illuminate underlying processes.
We believe copula approaches are among the tools all ecologists 
should be considering for analysis of their data in the 21st century.

@@

A third reason why we may expect an understanding of complex 
dependence structures and tail associations to benefit 
ecology is that such an approach may help illuminate 
causal mechanisms between variables. If two species,
Sp1 and Sp2, are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest Sp2
is the dominant competitor. Other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on associations. However, 
examination of tail associations may
be combined with biological reasoning to rule out
some causal hypotheses which could not previously be eliminated. 
@Popovic2019 used copulas to illuminate causal relationships
between species.

@@

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do positively associated 
ecological variables show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from a standard or default case. 
(Q2) If the answer to
Q1 is "yes", then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3), what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After section \ref{Background}, 'Background on copulas', we address Q1 
(sections \ref{Data} through \ref{Results} of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sections \ref{Causes} through \ref{Results_Q2}) using 
simple models. We address Q3 (sections \ref{Consequences} through \ref{Results_Q3}) using 
both data and models. Section \ref{Discussion} is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagrammatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have also been used effectively, but rarely so far, in ecology [@Valpine2014; 
@Anderson2018; @Popovic2019]. 
We argue that benefits of wider use of
copula descriptions of dependence in ecology will be substantial.

@@

We introduce some standard copula families, as examples, and also
because we will fit these families to data. The bivariate normal
copula family, which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of association
between the variables. A normal copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cdf of a (univariate) standard 
normal and $\Phi_{2,p}$ is the cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. Formulas for all the 
copula families we use were provided by @Brechmann2013. 
However, formulas for copulas are often not as 
directly informative about copula properties as probability density function (pdf) 
pictures; Fig. \ref{figped_N} has 
example pdfs corresponding to bivariate normal copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
associations between the two variables in the left and right tails. 
The Clayton copula family,
of which we already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to normal copulas, Clayton copulas
have stronger left- than right-tail association. The formula is 
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). Higher values of the parameter,
$p$ produce copulas with higher Kendall or Spearman correlation. The survival
Clayton family is the symmetric opposite of the Clayton family, showing
stronger right- than left-tail association (Fig. \ref{SM-figped_SC}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail association, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula, @joe2014_dependence and 
@Brechmann2013 for further
information on these and other families.

@@

Plots using normalized ranks were proposed in the Introduction as being conceptually
similar to copulas; we explain the connection. As stated above, the copula associated
with a random variable $(X,Y)$ is the cdf of the random variable
$(F_X(X),F_Y(Y))$. Given a sample $(x_i,y_i)$, $i=1,\ldots,n$ from $(X,Y)$, let $\hat{F}_X$ and
$\hat{F}_Y$ be the empirical cdfs associated with the $x_i$ and $y_i$, respectively.
These are step functions of $x$ and $y$, respectively, that start at $0$ for low values of 
$x$ and $y$ and jump by $1/n$
at each of the data points. Therefore $\hat{F}_X(x_j)$ is the rank of $x_j$ in
the set $\{x_i : i=1,\ldots,n\}$, divided by $n$, 
and analogously $\hat{F}_Y(y_j)$ is the rank of $y_j$ in the set 
$\{y_i : i=1,\dots,n\}$, divided
by $n$. For large $n$, $\hat{F}_X(x_i)$ and $\hat{F}_Y(y_i)$ 
approximate the cdfs $F_X$ and $F_Y$, respectively. 
But $\hat{F}_X(x_i)$ equals the normalized rank of $x_i$ times
$n/(n+1)$, and likewise $\hat{F}_Y(y_i)$ equals the normalized rank of $y_i$ times
$n/(n+1)$. So the normalized rank in turn approximates the empirical cdf and therefore 
the cdf. Thus the normalized rank pairs $(u_i,v_i)$ for $i=1,\ldots,n$ can be 
regarded as approximate samples from the random variable $(F_X(X),F_Y(Y))$, 
which is the random variable associated with the copula of $(X,Y)$. The scatterplot 
of $v_i$ versus $u_i$ can be used to infer aspects of copula 
structure (see section \ref{Methods} below).

@@

The datasets we used included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data (Table \ref{tab_data_info}), 
selected to span multiple fields and levels
of organization within ecology. Copula structure 
between atmospheric weather variables such as rainfall or wind speed, 
as measured in multiple locations through time, has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [RaCA; @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (megagrams C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for locations).
Species-trait data were average species basal metabolic rate (BMR, KJ per hour) and 
body mass (grams) for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the association has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years. 
Our second population dataset comprised average annual plankton abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells.
These data were computed from the Continuous Plankton Recorder survey of the 
UK Marine Biological Association.
Community-level data, obtained from the Cedar Creek Ecosystem Science Reserve, were 
plant aboveground biomass [@ccdataple120] and Shannon's diversity index (computed from plant 
species percent cover data [@ccdatapce120]) for `r nrow(d_cc[[5]])` plots [@ccdataplotse120], 
each 9m by 9m, as sampled in the years 1996-2000 and 2007, each year 
analyzed separately [@Tilman2001; @Tilman2006]. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. See Appendix \ref{SM-Data_details} for additional data details.

@@

<!--what we did and why is it important to study?-->
We showed that non-normal copula structure and asymmetric tail associations are 
common across multiple sub-disciplines
in ecology, although these facets of data are only occasionally 
accounted for [@Valpine2014; @Anderson2018; @Popovic2019]. We hypothesized 
mechanisms that may cause non-normal copula structures and asymmetric tail associations;
we discuss below how commonly some of our
mechanisms may operate. 
We also demonstrated important consequences of 
non-normal copula structure and asymmetric tail associations for ecology. 
For instance, the skewness
of a spatial-average time series is influenced by asymmetric tail associations between its 
constituent time series: predominantly right-tail-associated local time series 
can lead to "spiky" spatially averaged time series, with large outbreaks; and 
predominantly left-tail-associated local time series can lead to spatially
averaged time series showing accentuated "crashes".
Thus tail associations could have implications for pests and exploited species.
Extinction risk and Taylor's law can also be altered by tail association 
patterns across space. 
In our view, our results make it reasonable to suggest 
that a more comprehensive understanding 
of many ecological phenomena may be possible if a
complete copula characterization of associations between variables is employed. 
Copula statistics are
well developed [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017], 
and have been introduced in accessible formats [@Anderson2018; @Genest2007]. Furthermore, 
open-source computer implementations exist (e.g., the `copula` and 
`VineCopula` packages in R). Ecologists can apply these tools
immediately. We created several interrelated randomization procedures
(Appendices \ref{SM-surrog_test}, \ref{SM-Surrog_multi}, \ref{SM-sect:copsurrognd}) 
that built upon existing copula methods.

@@

<!-- We have only given a few mechanisms, but there are likely to be others, and discoverig
additional mechanisms of copula structure is certainly worthy of future work. -->
Additional mechanisms of non-normal copula structure and asymmetric tail associations
probably also operate. For instance, 
measurement error may modify copula structure. 
Our models investigating potential causes of copula structure 
were intentionally simple in other respects, 
too, not including factors such as
delayed density dependence, dispersal, population stage structure, trophic
interactions, etc.; and we did not comprehensively 
explore parameter space for our models. We re-emphasize 
that fuller explorations, in future work,
of some of our models and of variant models may be informative. 
We hope by enumerating a few potential mechanisms of copula structure
we will inspire additional research on the
potentially numerous mechanisms that may operate in diverse datasets,
and their relative importance under different circumstances.
